---
title: "DAS2021-Group-19"
author: "Xinyi Gao, Yiyang Li, Damien MacFarland, Neha Sinha, Jinda Zhang"
output: 
  pdf_document:
    number_sections: yes
fig_caption: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, eval=TRUE,warning = FALSE,message = FALSE, comment = NA)
library(ggplot2)
library(dplyr)
library(sjPlot)
library(stats)
library(janitor)
library(jtools)
library(moderndive)
library(gapminder)
library(skimr)
library(kableExtra)
library(gridExtra)
library(GGally)
library(MASS)
```

# Introduction {#sec:intro}

## Dataset 19

Dataset 19 comes from the Dallas animal shelter. You will have access to the following
variables, recorded by animal admission:

* `Animal_type` – The type of animal admitted to the shelter
* `Month` – Month the animal was admitted, recorded numerically with January=1
* `Year` – Year the animal was admitted to the shelter.
* `Intake_type` – Reason for the animal being admitted to the shelter
* `Outcome_type` – Final outcome for the admitted animal
* `Chip_Status` – Did the animal have a microchip with owner information?
* `Time_at_Shelter` – Days spent at the shelter between being admitted and the final outcome.


## Task

Imagine you have been asked by the shelter management to investigate the following questions of
interest:

* Which factors influence the number of days an animal spends in the shelter before their
final outcome is decided?

You should conduct an analysis to answer your question using a Generalised Linear Model (GLM).
Following your analyses, you should then summarise your results in the form of a presentation.

# Exploratory Data Analysis {#sec:EDA}
## Tidying Data

```{r upload csv}
shelter <- read.csv("dataset19.csv")
glimpse(shelter)
```
```{r tidying data}
shelter <- shelter %>% 
  mutate(animal=as.factor(animal_type),intake=as.factor(intake_type),outcome=as.factor(outcome_type),
         chip=as.factor(chip_status),time = time_at_shelter) %>% 
  dplyr::select(time,animal,intake,outcome,chip,month,year)

shelterbackup <- shelter
```

```{r visulaize time}
qntile <- quantile(shelter$time, 0.9)

ggplot(data=shelter, aes(x=time))+
  geom_histogram(fill="steelblue3",col="white", bins=30)+
  geom_vline(xintercept = qntile,color="red", linetype="dashed")+
  geom_text(aes(x=qntile+1,y=200,label="90 Percentile"),colour="red",angle=90,text=element_text(size=11))
```

```{r summary of time}
shelter %>%  dplyr::select(time) %>%
  summarise(mean=mean(time),s.d=sd(time),min=min(time),Q1=quantile(time,0.25), med=median(time),
  Q3=quantile(time,0.75),max=max(time))
```

```{r summary, eval=FALSE}
skim(shelter)
```

The time spent in the shelter is skewed - most animals (90%) stay no longer than two weeks and very few after three. The majority seem leave the shelter within one week. We could speculate that there are factors (possibly categorical variables) that contribute to the time at the shelter being shorter, and therefore if these factors do not occur then the animal stays longer.

We see that 90% of animals' outcome is decided by `r qntile` days at the shelter - this could be our research: what makes animals stay beyond 14 days? 

If we consider staying longer than 14 days as a failure, then we can reduce our problem to a binary GLM. 

```{r add success/failure}
shelter <- shelter %>% 
  mutate(on_time = ifelse(time>14,0,1)) %>% 
  mutate(on_time=as.factor(on_time))

levels(shelter$on_time) <- c("no","yes")
```

There are only two birds in the data and 13 wildlife. Therefore the impact these animal types will have on the other animal types will be minimal as they do not even contribute to 0.01% of the data, so they can be removed. 

```{r minus 2 birds + wildlife and outliers}
shelter <- shelter %>% filter(!(animal=="BIRD" | animal=="WILDLIFE"))

shelter <- shelter %>%  filter(time<=45)
```

```{r ggpairsio}
ggpairs(shelter)
```
## Time response variable

```{r animal explore}
ggplot(data = shelter, aes(x=on_time,y=animal,group=animal))+
  geom_boxplot(aes(fill=animal))+
  theme(legend.position = "none")

ggplot(data = shelter, aes(x=time,group=animal))+
  geom_bar(aes(fill=animal),stat = "count")+
  facet_wrap(~animal)
```
It would seem  that there is a slight difference in how long dogs and cats spend at the shelter by looking at the histograms, but from the boxplots we can see that both animals have very similar distributions: the medians are in similar places, their ranges are the similar, both have long tails (skewed right) and lots of outliers. It could be possible that this variable will be omitted from the final model. 

```{r intake explore}
ggplot(data = shelter, aes(y=time, x=intake,group=intake))+
  geom_boxplot(aes(fill=intake))+
  theme(legend.position = "none")

ggplot(data = shelter, aes(x=time,group=intake))+
  geom_bar(aes(fill=intake),stat = "count")+
  facet_wrap(~intake)
```

Confiscated animals seems to stay longer at the shelter than those animals that are surrendered by their owner or found as strays. There is a clear separation of one group from the other two. Strays seem to stay slighlty longer than animals that are surrendered. 

```{r outcome explore}
ggplot(shelter, aes(x=outcome,y=time,group=outcome))+
  geom_boxplot(aes(fill=outcome))+
  theme(legend.position = "none")

ggplot(data = shelter, aes(x=time,group=outcome))+
  geom_bar(aes(fill=outcome),stat = "count")+
  facet_wrap(~outcome)
```
There is some variation across outome - animals returned to owner or euthanized tend to spend short times at the shelter. Those that are adopted tend to be at the shelter for a slightly longer time and the distribution of this subgroup is more normally distributed. There is less data for died and fostered to make an assumption. 

```{r chip explore}
ggplot(shelter, aes(x=chip,y=time,group=chip))+
  geom_boxplot(aes(fill=chip))+
  theme(legend.position = "none")

ggplot(data = shelter, aes(x=time,group=chip))+
  geom_bar(aes(fill=chip),stat = "count")+
  facet_wrap(~chip)
```
Again all the data is skewed to the right, meaning most types stay for a short period of time. However, it would seem that animals with a chip is a lot more skewed that those that are not. We could speculate that these animals are rescued early and if not then held onto for longer, thus this variable could be a good predictor if combined with outcome perhaps.

```{r month explore}
ggplot(shelter, aes(x=month,y=time,group=month))+
  geom_boxplot(aes(fill=month))+
  theme(legend.position = "none")

ggplot(data = shelter, aes(x=time))+
  geom_bar(aes(fill=chip),stat = "count")+
  facet_wrap(~month)
```

The distributions for month seem to all be similar. We do not envisage that this variable will have an impact on the time variable. Similarly the variable year, which only has 2 levels, may not be a good predictor of time. that means we could potentially have 4 categorical variables explaining one continuous variable - **possible loglinear model**.

```{r 4 pairs categories boxplots}
ggplot(data=shelter,aes(x=time,fill=chip))+
  geom_boxplot(aes(y=outcome))

ggplot(data=shelter,aes(x=time,fill=chip))+
  geom_boxplot(aes(y=intake))

ggplot(data=shelter,aes(x=time,fill=animal))+
  geom_boxplot(aes(y=outcome))

ggplot(data=shelter,aes(x=time,fill=animal))+
  geom_boxplot(aes(y=intake))
```
Looking at the above plots, we see some change in distributions across different combinations of categories. 

```{r ggpairs}
g1 <- shelter %>% dplyr::select(-c(month,year)) %>% ggpairs(aes(fill=animal))
g1$nrow <- 1
g1$yAxisLabels <- g1$yAxisLabels[1]
g1
```
## on_time response variable

```{r on time props}
ggplot(shelter,aes(x=on_time,group=animal))+
  geom_bar(aes(y=..prop..,fill=animal),stat="count", position = "dodge")

ggplot(shelter,aes(x=on_time,group=intake))+
  geom_bar(aes(y=..prop..,fill=intake),stat="count", position = "dodge")

ggplot(shelter,aes(x=on_time,group=outcome))+
  geom_bar(aes(y=..prop..,fill=outcome),stat="count", position = "dodge")

ggplot(shelter,aes(x=on_time,group=chip))+
  geom_bar(aes(y=..prop..,fill=chip),stat="count", position = "dodge")

ggplot(data=shelter, aes(x=on_time,group=month))+
  geom_bar(aes(y=..prop..,fill=month),position="dodge")
```
There does not seem to be much difference in proportions between the two types of animals.
There is some difference in proportion between confiscated and the other levels of intake.
There is some differences from one group to another in the variable outcome.
There does not seem to be much differences in proportions for chip. 
For month, the proportions seem fairly even - cannot see any trend that could relate to seasonal activity. 

```{r continguencies}
ct_int <- table(shelter$intake ,shelter$on_time)
ct_anml <- table(shelter$animal,shelter$on_time)
ct_outcome <- table(shelter$outcome,shelter$on_time)
ct_chip <- table(shelter$chip,shelter$on_time)
mosaicplot(ct_int)
mosaicplot(ct_anml)
mosaicplot(ct_outcome)
mosaicplot(ct_chip)
```

Very little difference in proportions for yes/no across different levels of each categorical explanatory - maybe do not go for log-odds model. 

# Formal Data Analysis {#sec:FDA}

## Binomial models for on_time

```{r model 1}
model.full <- glm(data=shelter,on_time~animal+intake+outcome+chip,
              family = binomial(link="logit"))
model1 <- glm(data=shelter, on_time ~ outcome,
              family=binomial(link="logit"))
model2 <- glm(data=shelter, on_time ~ intake,
              family=binomial(link="logit"))
model3 <- glm(data=shelter, on_time ~ outcome + intake,
              family=binomial(link="logit"))
model4 <- glm(data=shelter, on_time ~ chip + intake,
              family=binomial(link="logit"))
model5 <- glm(data=shelter, on_time ~ animal + intake,
              family=binomial(link="logit"))

summ(model.full) #for comparison
summ(model1) #insignif coeffs
summ(model2) #no insignificant coeeficients
summ(model3) #insignif
summ(model4) #insignif
summ(model5) #insignif
```

BINOMIAL MODEL:From the exploratory analysis, it seemed fitting to try models based on intake and/or outcome, and then build up from there. Having fitted some models, model1 seems to be the best, which is simply intake as the only explanatory variable for the log-odds.

## Poisson models for time

```{r poi models}
model.full.poi <- glm(data=shelter, time~ intake+ animal + chip + outcome,
                      family=poisson())
model6 <- glm(data=shelter, time~intake + outcome,
              family = poisson())

summary(model.full.poi)  # dev =9916.4 on df=1828
confint(model.full.poi) # no 0's
summary(model6) # dev =10030 on df = 1831
confint(model6) # no 0's
```

```{r goodness of fit}
X2_full.poi <- sum(resid(model.full.poi,type="pearson")^2)
X2_6 <- sum(resid(model6,type="pearson")^2)

X2_full.poi < qchisq(df=1828,p=0.95)  #false - not good fit
X2_6 < qchisq(df=1831,p=0.95)         #false - not good fit
```


POISSON MODEL: looking at model6, using it has a very high pearson residual which is higher than the chi-squared statistic meaning it is probably suffering from overdispersion. To deal with this could be to introduce a dispersion parameter and have a quasi-poisson model OR we go for a negative binomial model.

```{r quasi poi}
dp <- X2_6 / model6$df.residual  #6.9337
summary(model6, dispersion = dp)
drop1(model6,test="F")
```

Coefficients still significant (by F-tests) but deviance still above chi-squared.

```{r negative bin}
mod.nbfull <- glm.nb(data=shelter, time ~ intake + outcome + chip + animal)
mod.nb6 <- glm.nb(data=shelter, time ~ intake + outcome)

summary(mod.nbfull) #smaller deviance than poi full: 2155
summary(mod.nb6) # smaller deviance than poi: 2153.4
```


# Conclusions {#sec:Conc}

